---
title: "ProblemSet3"
author: "Awesome computational biologist in training"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Problem Set 3

Week III: Sequence alignment and genome assembly

### Question 1 

__Globally align the following two sequences, using a gap opening penalty of 0 and a gap extension of 1 (look at the documentation for information on how to do this (?pairwiseAlignment)__ (1 mark) 

```{r question1}
library(Biostrings) 
library(pwalign)
library(xtable) 

seq1 = "TGGTACACGAAGGCATAAGT"
seq2 = "CATGGCTGGTGCTCCTGCTTGGCATGAAGCTGAATAAGAA"

mapping <- diag(4)
dimnames(mapping) <- list(DNA_BASES, DNA_BASES)
mapping[1,] = c(1,-2,-1,-2)  
mapping[2,] = c(-2,1,-2,-1)  
mapping[3,] = c(-1,-2,1,-2)  
mapping[4,] = c(-2,-1,-2,1)  

pairwiseAlignment(seq1, seq2,
                  type="global",
                  substitutionMatrix=mapping, fuzzyMatrix=NULL, 
                  gapOpening=0, gapExtension=1)
```

__Repeat this altering the gapExtension penalty for different values. What happens to the resulting alignments? __ (2 marks)
```{r}
for (i in seq(0, 12, 1)){
  result= pairwiseAlignment(seq1, seq2, type="global", substitutionMatrix = mapping, gapOpening=0, gapExtension = i, scoreOnly = FALSE)
  print(paste("gapExtension penalty:", i))
  print(result)
}
```
As the gap extension penalty increases, the score decreases because there is a higher penalty for every gap added. 
Moreover, the number of indels from the read sequence to the reference sequence decreases, while the number of substitutions
from the read sequence to the reference sequence increases. This is because increasing the gap extension penalty, while holding other 
penalties constant, result in gaps being increasingly penalized relative to substitutions. 

### Question 2

__The sequence below is the sequence of the homeobox domain, which of the sequences in the file (prot_seq.fa) has a homeobox domain - remember to justify your answer, what parameters did you use and why?__ (3 marks )

Sequence 1 and 8 have a homeobox domain
```{r}
getwd() #get working directory 
setwd("../data/problem-set-3") #set working directory 
prot_seq= readAAStringSet("prot_seq.fa")
```
```{r}
prot_seq
```
```{r}
homeodomain_seq="RRRKRTAYTRYQLLELEKEFHFNRYLTRRRRIELAHSLNLTERHIKIWFQNRRMKWKKEN"
```

```{r}
data("BLOSUM62", package = "pwalign")

for (i in 1:length(prot_seq)) {
  result <- pairwiseAlignment(homeodomain_seq, prot_seq[[i]], type = "local", substitutionMatrix =
                                BLOSUM62, gapOpening = 0, gapExtension =12, scoreOnly=FALSE)
  print(paste("sequence being aligned to homeodomain sequence:", names(prot_seq[i])))
  print(result)
}
```
Sequence 8 has a homeobox domain. This is because the substring in sequence 8 that is most similar
to the homeobox domain sequence has the same length as the homeobox domain sequence, which is not 
observed for other sequences. Moreover, it has the highest alignment score of 274 amongst all the
sequences. I chose to do local alignment because finding the substring in X (substring in protein sequence) 
which is most similar to the substring in Y (homeobox domain) requires local alignment. 
I chose BLOSUM62 as the substitution matrix because it is based on looking at amino acid substitutions 
at highly conserved stretches of protein, and homeobox domains are highly conserved. Moreover, BLOSUM62 
is based on local alignments. I used a gap extension penalty of -12. This is because the gap penalty that
we use for DNA alignments is -4 in general. 1 gap in a protein sequence corresponds to 3 gaps in DNA sequence because
an amino acid is coded by a codon, thus -4 x 3 bases = -12.

### Question 3

__Write a function sequenceGenome that given a genome (i.e. chr22.fa), generates a set of random reads (i.e 10000) of a specific length (i.e. 50bp) __ (3 marks)

```{r}
getwd() #get working directory 
setwd("../data/problem-set-3") #set working directory 
chr22genome = readDNAStringSet("chr22.fa")
```

```{r}
#create for loop for 10 reads of length 4 from chr22genome
read_vector= c()
for (i in 1:10){
  start= sample (1:width(chr22genome)-(4-1), size=1 , replace= TRUE)
  read<- as.character(extractAt(chr22genome, IRanges(start=start, end=start+4-1))[[1]])
  read_vector = c(read_vector, read)
  }
read_vector
```

```{r}
#build function 

sequenceGenome <- function(genome_name, read_length, number_of_reads){
  if (typeof(genome_name) == "S4" && class(read_length)== "numeric" && class(number_of_reads)== "numeric" && read_length<= width(genome_name)){
    read_vector= c()
    for (i in 1:number_of_reads){
      start= sample (1:width(genome_name)-(read_length-1), size=1 , replace= TRUE)
      read= as.character(extractAt(genome_name, IRanges(start=start, end=start+read_length-1))[[1]])
      read_vector = c(read_vector, read)
    }
    random_reads= DNAStringSet(read_vector)
    return(random_reads)}
  else { print ("Please enter the correct inputs: genome_name (DNAStringSet object of length 1), read_length (integer), number_of_reads (integer)")}
}
                              
random_reads= sequenceGenome(chr22genome, 10, 10)

random_reads
```

### Question 4 
__The function you made above did not allow for any variation in the reads - ( either due to sequencing error or due to variation) ... modify your function to randomly alter one base pair in each read to a different nucleotide (you will need to look for a function in Bioconductor/Biostrings that can alter one nucleotide)__ (1 marks)

```{r}
#randomly alter one base pair in one read to a different nucleotide

read= "ATGC" 
random_nucleotide_number= sample (1:nchar(read), size=1 , replace= TRUE) #randomly choose a number between 1 and the number of bases in the read
random_nucleotide= strsplit(read, "")[[1]][random_nucleotide_number] #choose a base that corresponds to the number above
change_to= sample (c("A", "G", "T", "C")[!c("A", "G", "T", "C")%in% c(random_nucleotide)], size=1 , replace= TRUE) #choose a random base that is not the base above
altered_read= replaceLetterAt(DNAString(read), c(random_nucleotide_number), change_to) #alter the read
altered_read #exhibit the altered read
```

```{r}

sequenceGenome_withErrors <- function(genome_name, read_length, number_of_reads){
  if (typeof(genome_name) == "S4" && class(read_length)== "numeric" && class(number_of_reads)== "numeric" && read_length<= width(genome_name)){
    read_vector= c()
    for (i in 1:number_of_reads){
      start= sample (1:width(genome_name)-(read_length-1), size=1 , replace= TRUE)
      read= as.character(extractAt(genome_name, IRanges(start=start, end=start+read_length-1))[[1]])
      random_nucleotide_number= sample (1:nchar(read), size=1 , replace= TRUE)
      random_nucleotide= strsplit(read, "")[[1]][random_nucleotide_number]
      change_to= sample (c("A", "G", "T", "C")[!c("A", "G", "T", "C")%in% c(random_nucleotide)], size=1 , replace= TRUE)
      altered_read= replaceLetterAt(DNAString(read), c(random_nucleotide_number), change_to)
      read_vector = c(read_vector, altered_read)
    }
    random_reads= DNAStringSet(read_vector)
    return(random_reads)}
  else { print ("Please enter the correct inputs: genome_name (DNAStringSet object of length 1), read_length (integer), number_of_reads (integer)")}
}

```

```{r}
sequenced_reads_with_errors = sequenceGenome_withErrors(chr22genome, 10, 20)
sequenced_reads_with_errors
```



```{r}
sessionInfo()
```
