---
title: "ProblemSet2"
author: "Awesome computational biologist in training"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Problem Set 1

Week II: Sequence alignment 

## Question 1

__Write a function (called hamming) that takes two strings and returns the Hamming distance between them__ (2 marks)
```{r question1}

#write a function (called hamming) that takes two strings and returns the Hamming distance between them
hamming <- function(seq1, seq2){
  distance<- 0 
  seq1<- strsplit(seq1, "")[[1]]
  seq2<- strsplit(seq2, "")[[1]]
  
  for (i in 1:length(seq1)){
    if (seq1[i]!= seq2[i]){
      distance<- distance + 1}
    else {distance <- distance}
  }
  print(distance)
}

#test the function to see if it returns the hamming distance

seq1 = "ATGC"
seq2 = "TTGC"
seq3 = "TTTC"

library(Biostrings)

hamming(seq1,seq2)
stringDist(c(seq1,seq2),method= "hamming")

hamming(seq1,seq3)
stringDist(c(seq1,seq3),method= "hamming")
```

If you want to make sure you're getting the right answer - remember you can use stringDist to check.

## Question 2

__Using this function, calculate the hamming distance between the proteins found in the file aa_seq.fa __ (1 marks)


```{r}
getwd() #get working directory 
setwd("../data/problem-set-2") #set working directory 
aa_seq= readAAStringSet("aa_seq.fa")

#length(aa_seq)
#as.character(aa_seq[1])
#as.character(aa_seq[2])
#hamming(as.character(aa_seq[1]),as.character(aa_seq[2]))

#create a matrix. the value in each cell represents the distance between two sequences.
#the sequence number corresponds to the row and column number of the cell. 
distance_matrix<- matrix(0, nrow = length(aa_seq), ncol = length(aa_seq))

#create a for loop to find the distance between the i-th and a-th sequence, for all i and a
for (i in 1: length(aa_seq)){
  for (a in 1: length(aa_seq)){
    if (i==a){
      distance_matrix[i,a] <- NA}
    else if (i!= a)
      {distance_matrix[i,a] <- hamming(as.character(aa_seq[i]), as.character(aa_seq[a]))}
  }}

#in this matrix, the value in each cell represents the distance between two sequences.
#the sequence number corresponds to the row and column number of the cell. 
distance_matrix 

#test if the hamming distance is correct

distance_matrix[3,5]
stringDist(c(aa_seq[3],aa_seq[5]),method= "hamming") 
```

__Given the hamming distances between these sequences which of the sequences is the most different (on average) to the other sequences__ (1 mark)
```{r}
#create a matrix
average_difference_matrix<- matrix(0, nrow = 1, ncol = length(aa_seq))

#store the mean distance in each column in a matrix
for (i in 1: length(aa_seq)){
    average_difference_matrix[1,i]<- mean(distance_matrix[,i], na.rm= TRUE)}

average_difference_matrix

which(average_difference_matrix== max(average_difference_matrix))
```
Sequence 1 is the most different (on average) to the other sequences

__Order the sequences in order of their similarity to seq5__ (1 mark)

```{r}
difference_df= data.frame(sequence= 1:10, difference_from_5= distance_matrix[5,])
difference_df <- difference_df[order(difference_df$difference_from_5, decreasing=TRUE),]

print(difference_df$sequence) #from least similar (sequence 7) to most similar (sequence 5)
```
From least to most similar to sequence 5: 7, 6, 1, 4, 8, 2, 3, 9, 10,  5

## Question 3 

__Read in the sequences in file ps2_sequences.fa, find what is the most abundant 6-mer for each of the sequences. Please make a function to do this for individual sequences.__ (4 marks)


```{r}
getwd() #get working directory 
setwd("../data/problem-set-2") #set working directory 
ps2_sequences=readDNAStringSet("ps2_sequences.fa")
ps2_sequences
#six_mers <- c(as.character(substring(ps2_sequences[1],1,6)))
#six_mers <- c(six_mers, as.character(substring(ps2_sequences[1],2,7)))

#ps2_sequences[1]

#six_mers<- c()
#for (i in 6: nchar(ps2_sequences[1])-5){
#  six_mers<- c(six_mers, as.character(substr(ps2_sequences[1],i,i+5)))
#}
#six_mers

most_abundant_sixmer<- function(sequence){
  six_mers<- c()
  for (i in 6: nchar(sequence)-5){
  six_mers<- c(six_mers, as.character(substr(sequence,i,i+5)))}
  names(which.max(table(six_mers)))
}

most_abundant_sixmer(ps2_sequences[1])

most_abundant_sixmer_vector<- c()
for (i in 1:length(ps2_sequences)){
  most_abundant_sixmer_vector<- c(most_abundant_sixmer_vector,most_abundant_sixmer(ps2_sequences[i]))
}

most_abundant_sixmer_df= data.frame(sequence= 1:5, most_abundant_sixmer= most_abundant_sixmer_vector)

most_abundant_sixmer_df
```

NB you can solve this is lots of different ways. 

__Given what the most abundant 6-mer in sequence 1 is ... what do you think this sequence might represent__ (1 mark).

```{r, eval=FALSE}

```

This might represent a telomeric sequence

```{r}
sessionInfo()
```
