---
title: "ProblemSet1"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Problem Set 1

Week I: Introduction to Computational Biology / Molecular Biology

## Question 1

__Read in the table of human chromosome sizes (hg38_sizes.txt), and plot a barplot of their sizes in Mb (megabases).__ (1 mark)
```{r question1}
hg38_sizes <- read.table("../data/problem-set-1/hg38_sizes.txt", quote="\"", comment.char="")
barplot(V2/1000000 ~ V1,
        data= hg38_sizes,
        las = 2,
        cex.names = 0.7,
        main = "Human Chromosome Sizes",
        xlab = "Chromosome",
        ylab = "Chromosome Size (Mb)",
        ylim= c(0,300))
```

Make sure you can read all of the chromosome names.

## Question 2

The file orf_coding.fasta contains the open reading frames in the yeast genome. Take the time to investigate what an open reading frame is 

__Plot a histogram of the size of the all of the orfs in the yeast genome and calculate the mean and median ORF size.__ (2 marks)
```{r question 2a}
library(Biostrings)
getwd() #get working directory #which folder is it accessing?
setwd("../data/problem-set-1") #set working directory #access this folder
yeast_orf= readDNAStringSet("orf_coding.fasta")

hist(nchar(yeast_orf),
     main = "Size of open reading frames in the yeast genome",
     col = "blue",
     ylim= c(0,2500),
     xlab= "Size (Mb)")

#mean ORF size
mean(nchar(yeast_orf))

#median ORF size
median(nchar(yeast_orf))
```
mean ORF size = 1485.306 Mb
median ORF size= 1218 Mb

__Do you think mean or median is a better way of representing the average ORF size (and why)?__ (1 mark)

```{r question 2b, eval=FALSE}

```

median is a better way of representing the average ORF size. this is because, as seen from the histogram, the distribution of ORF sizes is right-skewed, thus extremely large but rare ORF sizes will significantly affect the mean, but not the median.

### Question 3 

__Calculate the percentage of Gs and Cs for each ORF (this is called GC content) . How many ORFs have GC content greater than 50%?__ (2 marks)
```{r question3}
yeast_orf_ATGC<- letterFrequency(yeast_orf, c("A", "T", "G", "C"))

yeast_orf_GCcontent<- (yeast_orf_ATGC[,"G"]+ yeast_orf_ATGC[,"C"])/ (yeast_orf_ATGC[,"A"] + yeast_orf_ATGC[,"T"]+ yeast_orf_ATGC[,"G"]+ yeast_orf_ATGC[,"C"])*100

sum(yeast_orf_GCcontent>50)
```
78ORFs have GC content greater than 50%

### Question 4
__So all of these DNA sequences encode for potential proteins, transcribe all of these DNA sequences into RNA.__ (1 marks)
```{r question 4a}
RNAStringSet(yeast_orf) #transcribe DNA sequences into RNA
```

__So all of these DNA sequences encode for potential proteins, translate all of these DNA sequences into Protein.__ (1 mark)
```{r question 4b}
translate(yeast_orf) #translate DNA sequences into protein
```

### Question 5 
__Plot the distribution of amino acid usage of all yeast ORFs ... (i.e how many times is each individual amino acid encoded - you do not need to include STOP codons).__ (2 marks)

```{r question5}
yeast_aa_strings <- translate(yeast_orf) #store amino acid sequences of proteins

aa_list<- c("F","L","S","Y","C","W","P","H","Q","R","I","M","T","N","K","V","A","D","E","G") #store list of amino acids
aa_count_prot <- letterFrequency(yeast_aa_strings, aa_list) #create matrix to store the amino acid count of each protein
aa_count_prot<- as.data.frame(aa_count_prot) #convert matrix into dataframe

aa_count_total= c()
for( i in 1:ncol(aa_count_prot) ){
	  aa_count_total= c(aa_count_total,sum(aa_count_prot[,i]))
} #find the total number of each amino acid

barplot(aa_count_total,
        names.arg = aa_list,
        main= "Amino acid usage of all yeast ORFs",
        xlab= "Amino acid",
        ylab= "Frequency",
        ylim= c(0,300000),
        cex.names = 0.6)
```


```{r}
sessionInfo()
```
