---
title: "ProblemSet8"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Problem Set 8

For this problem set we will be working with the dataset that we have been analysing in the RNA-seq portion of this class. 

```{r}
library(DESeq2)
library(GenomicFeatures)
library(biomaRt)
```

```{r}
getwd() #get working directory 
setwd("../data/problem-set-8") #set working directory 

read_counts = read.delim("counts.txt", sep="\t", comment.char="#", row.names=1)

experimental.metadata = read.table("sampledata.txt", stringsAsFactors=FALSE, sep="\t", header=TRUE)

experimental.metadata$condition = factor(experimental.metadata$condition,
                             levels=c("mN","mP", "mT")) #restructure data
```
 
```{r}
getwd() #get working directory 
setwd("../data/shared") #set working directory 
mm.gtf.db <- makeTxDbFromGFF("Mus_musculus.GRCm39.109.chr.gtf", format="gtf" )


ensembl.genes = genes(mm.gtf.db)
gene_annotations <- rtracklayer::import("Mus_musculus.GRCm39.109.chr.gtf")
gene_annotations <- gene_annotations[gene_annotations$type == "gene"]
gene_metadata <- unique(as.data.frame(mcols(gene_annotations))[, c("gene_id", "gene_name", "gene_biotype")])
ensembl.genes$mgi_symbol = gene_metadata$gene_name[ match(ensembl.genes$gene_id, gene_metadata$gene_id) ]
ensembl.genes$gene_biotype = gene_metadata$gene_biotype[ match(ensembl.genes$gene_id, gene_metadata$gene_id) ]
ensembl.genes$status = NA
ensembl.genes$description = NA
ensembl.genes$entrezgene = NA
```
 
## Question 1:

### Question 1a:
__Lets play with TPMs. Load in the read_counts from this experiment and remove all rRNAs, mt_rRNAs, the gene Rn7sk and all of the genes that are located on the mitochondrial chromosome. We're doing this because these genes are sucking up a lot of reads in this dataset. Then recalculate the TPMs for all genes across all samples (1 mark)__ 

```{r}
genes_to_remove = names(ensembl.genes[ensembl.genes$mgi_symbol %in% "Rn7sk" | seqnames(ensembl.genes) == "chrM" | ensembl.genes$gene_biotype %in% c("Mt_rRNA", "rRNA" )])

read_counts = read_counts[ !(row.names(read_counts) %in% genes_to_remove),]
```

```{r}
ncscalingfactor = colSums(read_counts) / 1e6

mm.exons = exonsBy(mm.gtf.db, "gene")
mm.exons = reduce(mm.exons)
mm.lengths = sapply(mm.exons, function(x){sum(width(x))})

gene_lengths = mm.lengths / 1000 
rpk = read_counts / gene_lengths[row.names(read_counts)]
ncscalingfactor = colSums(rpk) / 1e6
tpm = sweep(rpk, 2, ncscalingfactor, "/")
```

```{r}
colSums(tpm)
```

```{r}
head(tpm)
```

### Question 1b:

__How many genes are expressed more than 1 TPM in at least one sample__ 
(1 mark)

```{r}
more_than_1_tpm= c()

for (rowname in rownames(tpm)){
  if (sum(tpm[rowname,] > 1) > 0 ){
    more_than_1_tpm = c(more_than_1_tpm, rowname)}
  else {}}
```

```{r}
print(paste(length(more_than_1_tpm), "genes are expressed more than 1 TPM in at least one sample"))
```

### Question 1c:

__Using only those genes which are expressed more than 1 TPM in at least one sample... for each sample identify which gene is closest to the median TPM for that sample.__ (1 mark)

```{r}
read_counts_more_than_1_tpm = read_counts[more_than_1_tpm,]
```

```{r}
for (colname in colnames(read_counts_more_than_1_tpm)){
  median_tpm <- median(read_counts_more_than_1_tpm[,colname])
  absolute_differences_from_median_tpm <- abs(read_counts_more_than_1_tpm[,colname] - median_tpm)
  minimum_absolute_difference_from_median_tpm <- min(absolute_differences_from_median_tpm)
  gene_closest_to_median_tpm <- rownames(read_counts_more_than_1_tpm[absolute_differences_from_median_tpm == minimum_absolute_difference_from_median_tpm,])
  
  print(paste("genes closest to the median TPM for sample", colname))
  print(gene_closest_to_median_tpm )
}
```

### Question 1d:

__Plot a boxplot showing the expression of Kras (in TPM) in the three conditions.__(1 mark)

```{r}
mN<- unlist(read_counts[names(ensembl.genes[ensembl.genes$mgi_symbol %in% "Kras",]),])[c("mN11","mN12","mN13","mN5","mN6",
                                                                                 "mN7")]
mT<- unlist(read_counts[names(ensembl.genes[ensembl.genes$mgi_symbol %in% "Kras",]),])[c("mT3", "mT4", "mT5", "mT7", "mT8", "mT9")]
mP<- unlist(read_counts[names(ensembl.genes[ensembl.genes$mgi_symbol %in% "Kras",]),])[c("mP1", "mP2", "mP3", "mP4", "mP5", "mP6")]

par(mfrow= c(1,3))
boxplot(mN, main= "Expression level of Kras in mN", ylim= c(1000,2400), ylab= "Expression level (TPM)")
grid()
boxplot(mT, main= "Expression level of Kras in mT", ylim= c(1000,2400))
grid()
boxplot(mP, main= "Expression level of Kras in mP", ylim= c(1000,2400))
grid()
```
## Question 2:

There are lots of ways to compare biological replicates.

### Question 2a:

__Write a loop that will generate pairwise scatterplots for all replicates within a condition, i.e. plot compare all replicates from the same condition against each other. By normalised counts I mean counts(dds, normalised=TRUE) .. one way is to plot the log10(normalised counts + 1) for one replicate against another.__ (1 marks)

```{r}
dds <- DESeqDataSetFromMatrix(read_counts, experimental.metadata, ~ condition)
dds <- estimateSizeFactors(dds) 
dds <- estimateDispersions(dds)
```
```{r}
mN_replicates <- c("mN11","mN12","mN13","mN5","mN6","mN7")

par(mfrow= c(3,2))
for (one_replicate in mN_replicates){
  replicate1 <- log10(counts(dds, normalized=TRUE) + 1)[,one_replicate]
  for (another_replicate in mN_replicates){
    replicate2 <- log10(counts(dds, normalized=TRUE) + 1)[,another_replicate]
    plot(replicate1, replicate2, main= paste(another_replicate, "against", one_replicate), ylab= "log10 (normalized counts + 1)", xlab= "log10 (normalized counts + 1)")}}
```

## Question 2b

__Instead of using log10(x+1), plot the counts generated using the rlog function (again comparing all replicates from a specific condition)?__(1 marks)

```{r}
rld <- rlog(dds)
rld <- assay(rld)

mN_replicates <- c("mN11","mN12","mN13","mN5","mN6","mN7")

par(mfrow= c(3,2))
for (one_replicate in mN_replicates){
  replicate1 <- rld[,one_replicate]
  for (another_replicate in mN_replicates){
    replicate2 <- rld[,another_replicate]
    plot(replicate1, replicate2, main= paste(another_replicate, "against", one_replicate), ylab= "normalized counts", xlab= "normalized counts")}}
```


__How are the two plots different .. and why?__ ( 1 marks )

By eye, there is way more heteroskedasticity (i.e. the plot thickens more) in the first plot, as compared to in the second plot. 

This is because rlog "transforms the count data to the log2 scale in a way which minimizes differences between samples for rows with small counts, and which normalizes with respect to library size", which is not the case with log10(x+1). 

## Question 2c

__MA plots are incredibly powerful in comparing samples against each other, using the results from counts(dds, normalized=TRUE) plot MA plots for each pair of replicates within a condition__(2 marks)

```{r}
dds <- DESeqDataSetFromMatrix(read_counts, experimental.metadata, ~ condition)
dds <- estimateSizeFactors(dds) 
dds <- estimateDispersions(dds)
```

```{r}
mN_replicates <- c("mN11","mN12","mN13","mN5","mN6","mN7")

par(mfrow= c(2,2))
for (one_replicate in mN_replicates){
  replicate1 <- counts(dds, normalized=TRUE)[,one_replicate] 
  for (another_replicate in mN_replicates){
    if (one_replicate != another_replicate){
    replicate2 <- counts(dds, normalized=TRUE)[,another_replicate] 
    M= log2(replicate1/ replicate2)
    A= 0.5*log2(replicate1 * replicate2)
    plot(A,
         M, 
         main= paste("MA plot:", one_replicate, "vs", another_replicate),
         ylab= paste("log2 (counts in", one_replicate," / counts in", another_replicate, ")"),
         xlab= paste("0.5 * log2 (counts in", one_replicate," * counts in", another_replicate, ")") 
         )
    }
    else{}
  }
  }
```


__What does an MA plot show? How does it work / what does it show / how can you interpret this?__(1 mark)

The MA plot is basically plotting log2(normalized counts in sample A / normalized counts in sample B)
against 0.5 * log2(normalized counts in sample A * normalized counts in sample B), where each point on the plot corresponds to a single gene. The more positive the y axis coordinate of a point is, the more highly the gene corresponding to it is expressed on average in the two samples A and B. The closer the x axis coordinate of a point is to 0, the less different the gene expression is between the two samples A and B. However, this difference does not refer to statistical significance. If the x axis coordinate of a point is positive, then the gene is more highly expressed in sample A relative to sample B. If the x axis coordinate of a point is negative, then the gene is expressed less in sample A relative to sample B. 

```{r}
sessionInfo()
```
